Compositional equivalences based on Open pNets

### 2020-07-21 2016-Weak-FH-Bisim/WeakBisimSUBMITTED.tex
submitted version

\title{Compositional equivalences based on Open pNets
%\thanks{This work was partially funded by the Associated Team FM4CPS
%  between INRIA and ECNU, Shanghai}
}
\author{Ludovic Henrio}
\address{Univ Lyon, EnsL, UCBL, CNRS, Inria,  LIP, F-69342, LYON Cedex 07, France.}
\email{ludovic.henrio@cnrs.fr}

\author{Eric Madelaine}
\address{INRIA Sophia Antipolis M\'edit\'erann\'ee, UCA, BP 93, 06902 Sophia Antipolis, France}
\email{eric.madelaine@inria.fr}
%\date{}                                           % Activate to display a given date or no date
\author{Rab\'ea Ameur-Boulifa}
\address{LTCI, T\'el\'ecom Paris, Institut Polytechnique de Paris, France
}         
\email{rabea.ameur-boulifa@telecom-paris.fr}                % Activate to display a given date or no date

@@

\begin{figure}[t]
   \includegraphics[width=.37\textwidth]{XFIG/SimpleProt-Schema}
   \includegraphics[width=.62\textwidth]{XFIG/SimpleProt2-Spec}
   \caption{pNet structure of the example and specification expressed as a pNet}
   \label{SimpleProt:Spec}

@@

\begin{figure}[t]
  \centerline{\includegraphics[width=.9\textwidth]{XFIG/SimpleProt2-pNet-tau}}
  \caption{Composed pNet with the Simple Protocol Implementation}  \label{SimpleProt:Impl}
\end{figure}

@@

 Then 
$\mathcal{R}$ is an FH-bisimulation if and only if for any  states
$s\in\mathcal{S}_1$ and $t\in\mathcal{S}_2$, $(s,t|\Pred_{s,t})\in\mathcal{R}$, we 
have
the following:
\end{minipage}
\hspace{2mm}
\begin{minipage}{0.30\linewidth}
	\includegraphics[width=\linewidth]{XFIG/Bisim}
\end{minipage}

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=13cm]{XFIG/SPSpecOpen}}
   \caption{Specification Open Automaton}
   \label{SimpleProtCounter:SpecOA}
\end{figure}

@@

 \begin{figure}[ht]
  \centerline{\includegraphics[width=15cm]{XFIG/SPImplOpen}}
  \caption{Open Automaton of the Simple Protocol Implementation}  \label{SimpleProtCounter:ImplOA}
\end{figure}

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=13cm]{XFIG/WOT2-result}}
  \caption{Construction of an example of weak open transition}
   \label{WOT2}
\end{figure}

@@

\begin{figure}[h]
   \centerline{\includegraphics[width=15cm]{XFIG/SPSpecWeakOpen}}
  \caption{Weak Open Automaton of the Specification}
   \label{SimpleProtCounter:WeakSpecOA}
\end{figure}

@@

%\begin{figure}[h]
  % \centerline{\includegraphics[width=11cm]{XFIG/SimpleProtImpl-WOA}}
 % \caption{Weak Open Automaton of the implementation}
 %  \label{SimpleProtCounter:ImplWOA}
%\end{figure}

\begin{figure}[h]
   \centerline{\includegraphics[width=11cm]{XFIG/SimpleProtImpl-WOA2}}
  \caption{Weak Open Automaton of the implementation}
   \label{SimpleProtCounter:ImplWOA2}
\end{figure}

@@

\ERIC{ To the best of our knowledge, there are not many research works on Weak Bisimulation Equivalences between such complicate system models (open, symbolic, data-aware, with loops and assignments).
We give a brief overview of other related publications, focussing first on Open and Compositional approaches, then on Symbolic Bisimulation for data-sensitive systems.}

@@

\subsubsection*{Open and Compositional systems}
%\TODO{retravailler le paragraphe ci dessous: faire ref a Facs-2013 (reconfig ds pNets) - virer 14 (probabilistic) ?} \cite{GasparHM:2013}
\ERIC{In
\cite{Johnson:2013:CBSE,Johnson:2014:qosa}, the authors investigate several methodologies for the compositional verification of software systems, with the aim to verify reconfigurable component systems. To improve scaling and compositionality, the authors  decouple the verification problem that is to be resolved by a SMT (satisfiability modulo theory) solver into independent sub-problems on independent sets of variables. These works clearly highlight the interest of incremental and compositional verification in a very general setting. In our own work on open pNets, adding more structure to the composition model, we show how to enforce a compositional proof system that is more powerful than independent sets of variables. Our theory has also been encoded into an SMT solver and it would be interesting to investigate how the examples of evolving systems studied by the authors could be encoded into pNet and verified by our framework. However, the models of Johnson et al. are quite different from ours, in particular they are much less structured, and translating them is clearly outside the scope of this article.
In previous work \cite{gaspar:hal-00916115}, we also have shown how (closed) pNet models could be used to encode and verify finite instances of reconfigurable component systems. }
%Additionally, the more compositional constraint resolution provided by the authors could also be interesting to make our SMT proofs scale better (inside a single pNet for example), the integration of the Z3 framework described in \cite{Johnson:2014:qosa} with our proof theory for open pNets would also be worth investigating from a practical point of view. 
%One major different however is that pNets in their current state are not much adapted to probabilistic verification contrarily to \cite{Johnson:2014:qosa} but they are better suited to verify complex behavioural properties.

@@

\subsubsection*{Symbolic and data-sensitive systems}
\ERIC{ As mentioned in the \emph{Introduction}, the work that brought us a lot of inspirations are those of Lin et al. \cite{IngolfsdottirL:2001,HennessyLin:TCS95,Linconcur96}. They developed the theory of symbolic transition graphs (STG), and the associated symbolic (early and late, strong and weak) bisimulations, they also study STGs with assignments as a model for message-passing processes. Our work extends these in several ways: first our models are compositional, and our bismulations come with effective conditions for being preserved by pNet composition (i.e. congruent), even for the weak version. This result is more general than the bisimulation congruences for value-passing CCS in \cite{IngolfsdottirL:2001}. Then our settings for management of data types are much less restrictive, thanks to our use of satisfiability engines, while Lin's algorithms were limited to data-independent systems. }

@@

We are currently extending this work,  looking at  both further properties of FH-bisimulation, but also
the relations with existing equivalences on both closed and open systems. \ERIC{In particular, our model being significantly different from those considered in \cite{IngolfsdottirL:2001}, it
would be interesting to compare our "FH" family of bisimulations with the hierarchy of symbolic bisimulations from these authors.}
We also plan to apply open pNets to the study of complex composition
operators in a symbolic way, for example in the area of parallel
skeletons, or distributed algorithms.
We have developed tool support for computing the
symbolic semantics in term of open-automata \cite{QBMZ-AVOCS18}, and have
developed algorithms to check strong FH-bisimulation \cite{hou:hal-02406098}.  
Naturally we are now working toward an implementation of weak FH-Bisimulation. 
The challenges here, in the context of our symbolic systems, is not so much algorithmic complexity,
as was the case with classical weak bisimulation on finite models, but decidability and termination.
Two main directions should be explored, either with an explicit construction of the weak transition, but this step in
itself may introduce non-termination, or a direct implementation of the weak bisimulation definition, without
constructing the weak automaton. We can expect that optimisations here in the symbolic setting would be quite different from the classical case.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{XFIG/proof-explain}
\end{center}\caption{Composition of the subsequences}\label{figproof}
\end{figure}

@@

Note also that $\Post$ only acts on variables of $P$ while $\Post_{1i}$ only acts on  variables of $Q$. We conclude on predicate and posts as follows\footnote{$\Post_{1i}$ only has an effect on variables of $Q$ and thus does not modify $\Pred$ or $\beta_{j_0}$}:
\begin{equation*}
\begin{split}
\Pred_c&=\Pred_Q\land\Pred\subst{\bigotimes_{i=k}^{0}{\Post_{1i}}}\land
(\alpha'_Q=\beta_{j_0})\subst{\bigotimes_{i=k}^{0}{\Post_{1i}}}\\
&=\Pred_Q\land\Pred\land \alpha_Q=\beta_{j_0}\\[.3ex]
\Post_c&=\Post\uplus\Post_Q \qedhere
\end{split}
\end{equation*}
\end{proof}
\smallskip
\begin{lemma}[Weak open transition composition]\label{lem-Weakcompose} Suppose that we have a weak open automaton such that the WOTs cannot observe silent actions (see Definition \ref{def:Non-ObsTau}). Suppose $j_0\in J$ and $\gamma_{j_0}=[\beta_0..\beta_n]$ and additionally:\\[-2ex]
\begin{mathpar}
%\[
P\models{\openrule
	{
		\gamma_j^{j\in J}, 
		\Pred,  
		\Post}
	{\ostate{s_i^{i\in L}} \OTWeakarrow {\alpha}
		\ostate{s_i'^{\, i\in L}}}
}
\qquad\text{~~and for all $p\in[0..n]$~~}\quad
Q\models{\openrule
	{
		\gamma_{pj}^{j\in J_p},
		 \Pred_p,  
		\Post_p}
	{\ostate{s_{pi}^{i\in L_Q}} \OTWeakarrow {\alpha_p}
		\ostate{s_{(p+1) i}^{\, i\in L_Q}}}
}%\]
\end{mathpar}

@@

%% \begin{figure}[t]
%%    \includegraphics[width=5.5cm]{XFIG/SimpleProt-Schema}
%%    \includegraphics[width=6.5cm]{XFIG/SimpleProt2-Spec}
%%    \caption{Specification pNet Schema and Specification}
%%    \label{App:SimpleProt:Spec}

@@

%% \begin{figure}[t]
%%   \centerline{\includegraphics[width=12cm]{XFIG/SimpleProt2-pNet}}
%%   \caption{Composed pNet with the Simple Protocol Implementation}  \label{App:SimpleProt:Impl}
%% \end{figure}

@@

\begin{figure}[h]
   \centerline{\includegraphics[width=10cm]{XFIG/SimpleProtImpl-WOA2}}
   \caption{Weak Open Automaton of the implementation}
   \label{Appendix:ImplOA2}
 \end{figure}

>>>>

\title{Compositional equivalences based on Open pNets
%\thanks{This work was partially funded by the Associated Team FM4CPS
%  between INRIA and ECNU, Shanghai}
}
\author{Rab\'ea Ameur-Boulifa}
\address{LTCI, T\'el\'ecom Paris, Institut Polytechnique de Paris, France
}         
\email{rabea.ameur-boulifa@telecom-paris.fr}

\author{Ludovic Henrio}
\address{Univ Lyon, EnsL, UCBL, CNRS, Inria,  LIP, F-69342, LYON Cedex 07, France.}
\email{ludovic.henrio@cnrs.fr}

\author{Eric Madelaine}
\address{INRIA Sophia Antipolis M\'edit\'erann\'ee, UCA, BP 93, 06902 Sophia Antipolis, France}
\email{eric.madelaine@inria.fr}
%\date{}                                           % Activate to display a given date or no date
               % Activate to display a given date or no date

@@

\begin{figure}[t]
   \includegraphics[width=.37\textwidth]{SimpleProt-Schema.pdf}
   \includegraphics[width=.62\textwidth]{SimpleProt2-Spec.pdf}
   \caption{pNet structure of the example and specification expressed as a pNet}
   \label{SimpleProt:Spec}

@@

\begin{figure}[t]
  \centerline{\includegraphics[width=.9\textwidth]{SimpleProt2-pNet-tau.pdf}}
  \caption{Composed pNet with the Simple Protocol Implementation}  \label{SimpleProt:Impl}
\end{figure}

@@

 Then 
$\mathcal{R}$ is an FH-bisimulation if and only if for any  states
$s\in\mathcal{S}_1$ and $t\in\mathcal{S}_2$, $(s,t|\Pred_{s,t})\in\mathcal{R}$, we 
have
the following:
\end{minipage}
\hspace{2mm}
\begin{minipage}{0.30\linewidth}
	\includegraphics[width=\linewidth]{Bisim.pdf}
\end{minipage}

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=13cm]{SPSpecOpen.pdf}}
   \caption{Specification Open Automaton}
   \label{SimpleProtCounter:SpecOA}
\end{figure}

@@

 \begin{figure}[ht]
  \centerline{\includegraphics[width=15cm]{SPImplOpen.pdf}}
  \caption{Open Automaton of the Simple Protocol Implementation}  \label{SimpleProtCounter:ImplOA}
\end{figure}

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=13cm]{WOT2-result.pdf}}
  \caption{Construction of an example of weak open transition}
   \label{WOT2}
\end{figure}

@@

\begin{figure}[h]
   \centerline{\includegraphics[width=15cm]{SPSpecWeakOpen.pdf}}
  \caption{Weak Open Automaton of the Specification}
   \label{SimpleProtCounter:WeakSpecOA}
\end{figure}

@@

%\begin{figure}[h]
  % \centerline{\includegraphics[width=11cm]{SimpleProtImpl-WOA}}
 % \caption{Weak Open Automaton of the implementation}
 %  \label{SimpleProtCounter:ImplWOA}
%\end{figure}

\begin{figure}[h]
   \centerline{\includegraphics[width=11cm]{SimpleProtImpl-WOA2.pdf}}
  \caption{Weak Open Automaton of the implementation}
   \label{SimpleProtCounter:ImplWOA2}
\end{figure}

@@

 To the best of our knowledge, there are not many research works on Weak Bisimulation Equivalences between such complicate system models (open, symbolic, data-aware, with loops and assignments).
We give a brief overview of other related publications, focussing first on Open and Compositional approaches, then on Symbolic Bisimulation for data-sensitive systems.

@@

\subsubsection*{Open and Compositional systems}
%\TODO{retravailler le paragraphe ci dessous: faire ref a Facs-2013 (reconfig ds pNets) - virer 14 (probabilistic) ?} \cite{GasparHM:2013}
In
\cite{Johnson:2013:CBSE,Johnson:2014:qosa}, the authors investigate several methodologies for the compositional verification of software systems, with the aim to verify reconfigurable component systems. To improve scaling and compositionality, the authors  decouple the verification problem that is to be resolved by a SMT (satisfiability modulo theory) solver into independent sub-problems on independent sets of variables. These works clearly highlight the interest of incremental and compositional verification in a very general setting. In our own work on open pNets, adding more structure to the composition model, we show how to enforce a compositional proof system that is more powerful than independent sets of variables. Our theory has also been encoded into an SMT solver and it would be interesting to investigate how the examples of evolving systems studied by the authors could be encoded into pNet and verified by our framework. However, the models of Johnson et al. are quite different from ours, in particular they are much less structured, and translating them is clearly outside the scope of this article.
In previous work \cite{gaspar:hal-00916115}, we also have shown how (closed) pNet models could be used to encode and verify finite instances of reconfigurable component systems. 
%Additionally, the more compositional constraint resolution provided by the authors could also be interesting to make our SMT proofs scale better (inside a single pNet for example), the integration of the Z3 framework described in \cite{Johnson:2014:qosa} with our proof theory for open pNets would also be worth investigating from a practical point of view. 
%One major different however is that pNets in their current state are not much adapted to probabilistic verification contrarily to \cite{Johnson:2014:qosa} but they are better suited to verify complex behavioural properties.

@@

\subsubsection*{Symbolic and data-sensitive systems}
 As mentioned in the \emph{Introduction}, the work that brought us a lot of inspirations are those of Lin et al. \cite{IngolfsdottirL:2001,HennessyLin:TCS95,Linconcur96}. They developed the theory of symbolic transition graphs (STG), and the associated symbolic (early and late, strong and weak) bisimulations, they also study STGs with assignments as a model for message-passing processes. Our work extends these in several ways: first our models are compositional, and our bismulations come with effective conditions for being preserved by pNet composition (i.e. congruent), even for the weak version. This result is more general than the bisimulation congruences for value-passing CCS in \cite{IngolfsdottirL:2001}. Then our settings for management of data types are much less restrictive, thanks to our use of satisfiability engines, while Lin's algorithms were limited to data-independent systems.

@@

We are currently extending this work,  looking at  both further properties of FH-bisimulation, but also
the relations with existing equivalences on both closed and open systems. In particular, our model being significantly different from those considered in \cite{IngolfsdottirL:2001}, it
would be interesting to compare our "FH" family of bisimulations with the hierarchy of symbolic bisimulations from these authors.
We also plan to apply open pNets to the study of complex composition
operators in a symbolic way, for example in the area of parallel
skeletons, or distributed algorithms.
We have developed tool support for computing the
symbolic semantics in term of open-automata \cite{QBMZ-AVOCS18}, and have
developed algorithms to check strong FH-bisimulation \cite{hou:hal-02406098}.  
Naturally we are now working toward an implementation of weak FH-Bisimulation. 
The challenges here, in the context of our symbolic systems, is not so much algorithmic complexity,
as was the case with classical weak bisimulation on finite models, but decidability and termination.
Two main directions should be explored, either with an explicit construction of the weak transition, but this step in
itself may introduce non-termination, or a direct implementation of the weak bisimulation definition, without
constructing the weak automaton. We can expect that optimisations here in the symbolic setting would be quite different from the classical case.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{proof-explain.pdf}
\end{center}\caption{Composition of the subsequences}\label{figproof}
\end{figure}

@@

Note also that $\Post$ only acts on variables of $P$ while $\Post_{1i}$ only acts on  variables of $Q$. We conclude on predicate and posts as follows\footnote{$\Post_{1i}$ only has an effect on variables of $Q$ and thus does not modify $\Pred$ or $\beta_{j_0}$}:
\begin{equation*}
\begin{split}
\Pred_c&=\Pred_Q\land\Pred\subst{\bigotimes_{i=k}^{0}{\Post_{1i}}}\land
(\alpha'_Q=\beta_{j_0})\subst{\bigotimes_{i=k}^{0}{\Post_{1i}}}\\
&=\Pred_Q\land\Pred\land \alpha_Q=\beta_{j_0}\\[.3ex]
\Post_c&=\Post\uplus\Post_Q 
\end{split}
\end{equation*}
\end{proof}
\smallskip
\begin{lemma}[Weak open transition composition]\label{lem-Weakcompose} Suppose that we have a weak open automaton such that the WOTs cannot observe silent actions (see Definition \ref{def:Non-ObsTau}). Suppose $j_0\in J$ and $\gamma_{j_0}=[\beta_0..\beta_n]$ and additionally:\\[-2ex]
\begin{mathpar}
%\[
P\models{\openrule
	{
		\gamma_j^{j\in J}, 
		\Pred,  
		\Post}
	{\ostate{s_i^{i\in L}} \OTWeakarrow {\alpha}
		\ostate{s_i'^{\, i\in L}}}
}
\qquad\text{~~and for all $p\in[0..n]$~~}\quad
Q\models{\openrule
	{
		\gamma_{pj}^{j\in J_p},
		 \Pred_p,  
		\Post_p}
	{\ostate{s_{pi}^{i\in L_Q}} \OTWeakarrow {\alpha_p}
		\ostate{s_{(p+1) i}^{\, i\in L_Q}}}
}%\]
\end{mathpar}

@@

%% \begin{figure}[t]
%%    \includegraphics[width=5.5cm]{SimpleProt-Schema}
%%    \includegraphics[width=6.5cm]{SimpleProt2-Spec}
%%    \caption{Specification pNet Schema and Specification}
%%    \label{App:SimpleProt:Spec}

@@

%% \begin{figure}[t]
%%   \centerline{\includegraphics[width=12cm]{SimpleProt2-pNet}}
%%   \caption{Composed pNet with the Simple Protocol Implementation}  \label{App:SimpleProt:Impl}
%% \end{figure}

@@

\begin{figure}[h]
   \centerline{\includegraphics[width=10cm]{SimpleProtImpl-WOA2.pdf}}
   \caption{Weak Open Automaton of the implementation}
   \label{Appendix:ImplOA2}
 \end{figure}
